---
title: "Clone Selection"
author: 
  - Costa, W. G.^[Weverton Gomes da Costa, Pós-Doutorando, Embrapa Mandioca e Fruticultura, wevertonufv@gmail.com]
date: "`r Sys.Date()`"
site: workflowr::wflow_site
url: https://wevertongomescosta.github.io/Genomic-Selection-for-Drought-Tolerance-Using-Genome-Wide-SNPs-in-Casava
output:
  workflowr::wflow_html:
    toc: TRUE
editor_options:
  chunk_output_type: console
---

## Configurations and packages

```{r setup, include=F} 
knitr::opts_chunk$set(echo = T, warning = F)
```

To perform the analyses, we will need the following packages:

```{r libraries, message=FALSE}
library(tidyverse)
library(kableExtra)
library(ggthemes)
library(metan)
library(psych)
library(viridis)
library(ggpubr)
```

## Data

```{r data}
results <- readRDS("output/results_cv_G_BLUP.RDS") %>%
  mutate(method = "G-BLUP") %>%
  bind_rows(
    readRDS("output/results_cv_RR_BLUP.RDS") %>%
      mutate(method = "RR-BLUP"),
    readRDS("output/results_cv_RKHS.RDS") %>%
      mutate(method = "RKHS"),
    readRDS("output/results_cv_BayesA.RDS") %>%
      mutate(method = "Bayes A"),
    readRDS("output/results_cv_BayesB.RDS") %>%
      mutate(method = "Bayes B"),
    readRDS("output/results_cv_RF.RDS") %>%
      mutate(method = "RF"),
    readRDS("output/results_cv_GEBVS_DOM.RDS") %>%
      mutate(method = "G-BLUP-DOM")
  )

traits <- unique(results$Trait)
```

## Results

### Plot boxplot from Accuracy

Figure 2 Boxplot of predictive ability 

```{r accuracy}
results %>%
  ggplot(aes(x = method, y = Ac, fill = method)) +
  geom_boxplot() +
  facet_wrap(~ Trait, ncol = 6) +
  expand_limits(y = 0) +
  labs(y = "Accuracy", x = "", fill = "Method") +
  scale_fill_gdocs() +
  theme(
    text = element_text(size = 25),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    legend.position = "top",
    legend.title = element_blank(),
    legend.box = "horizontal",
    panel.spacing = unit(1, "lines"),
    strip.background = element_blank(),
    panel.background = element_blank(),
    plot.background = element_blank(),
    legend.background = element_blank(),
    legend.box.background = element_blank(),
    legend.key = element_blank()
  ) +
  guides(fill = guide_legend(
    nrow = 1,
    byrow = TRUE,
    keywidth = 1.5,
    keyheight = 1,
    title.position = "top"
  ))
```


```{r accuracy_save, include=FALSE}
ggsave(
  "output/accuracy_all_methods.tiff",
  width = 16,
  height = 10,
  dpi = 300
)
```

Table to Figure 2 Boxplot of predictive ability 

```{r accuracy_table}
# Calcular médias e desvio padrão de Ac por Trait e método
results_Ac <- results %>%
  group_by(Trait, method) %>%
  summarise(
    Ac_mean = round(mean(Ac) * 100, 2),
    Ac_sd = round(sd(Ac) * 100, 2),
    .groups = "drop"  # Remove agrupamento após summarise
  ) %>%
  select(Trait, method, Ac_mean) %>%
  pivot_wider(names_from = method, values_from = Ac_mean)

# Exibir resultados em tabela com kable
results_Ac %>%
  kbl(escape = FALSE, align = "c") %>%
  kable_classic(
    "hover",
    full_width = FALSE,
    position = "center",
    fixed_thead = TRUE
  )
```

```{r MSPE_table}
# Calcular médias e desvio padrão de MSPE por Trait e método
results_MSPE <- results %>%
  group_by(Trait, method) %>%
  summarise(
    MSPE_mean = round(mean(MSPE) * 100, 2),
    MSPE_sd = round(sd(MSPE) * 100, 2),
    .groups = "drop"  # Remove agrupamento após summarise
  ) %>%
  select(Trait, method, MSPE_mean) %>%
  pivot_wider(names_from = method, values_from = MSPE_mean)

# Exibir resultados em tabela com kable
results_MSPE %>%
  kbl(escape = FALSE, align = "c") %>%
  kable_classic(
    "hover",
    full_width = FALSE,
    position = "center",
    fixed_thead = TRUE
  )
```

```{r save_results, include=FALSE}
write.csv(results_Ac, "output/results_accuracy.csv", row.names = FALSE)

write.csv(results_MSPE, "output/results_MSPE.csv", row.names = FALSE)
```

### Clone Selection

First let's add the phenotypic means to the BLUPS and GEBVS

```{r clone_selection}
# Carregar dados
media_pheno <- read.csv("output/mean_pheno.csv")

BLUPS <- readRDS("data/pheno.rds") %>%
  pivot_longer(cols = -ID_Clone, names_to = "Trait", values_to = "BLUP")

# Inserir o método em cada data frame dentro de results$result
results$result <- map2(results$result, results$method, ~mutate(.x, method = .y))

# Combinar todos os data frames em um único data frame
GEBV_BLUP <- bind_rows(results$result) %>%
  group_by(ID_Clone, Trait, method) %>%
  summarise(GEBV = mean(GEBV), .groups = "drop") %>%
  pivot_wider(names_from = method, values_from = GEBV) %>%
  full_join(BLUPS, by = c("ID_Clone", "Trait"))

# Adicionar as médias fenotípicas aos valores numéricos e combinar com GEBV
GEBV_BLUP <- GEBV_BLUP %>%
  rowwise() %>%
  mutate(across(where(is.numeric), ~ . + media_pheno[[Trait]])) %>%
  ungroup()

# Visualizar os primeiros dados
GEBV_BLUP %>%
  head() %>%
  kbl(escape = FALSE, align = "c") %>%
  kable_classic("hover", full_width = FALSE, position = "center", fixed_thead = TRUE)
```


```{r clone_selection_save, include=FALSE}
write.csv(GEBV_BLUP, file ="output/GEBV_BLUP.csv" , row.names = FALSE)
```

Now let's group the BLUPs data with the GEBVs and GETGVs data and add a Weights column for each increase or decrease characteristic.

```{r pesos} 
selection_parents <- GEBV_BLUP %>% 
  rename(GEBV = `G-BLUP`, GETGV = `G-BLUP-DOM`) |>
  mutate(Weights = ifelse(
    Trait %in% traits,
    "acrescimo",
    "descrescimo"
  ))

calcular_pesos <- function(data, var){
  selection_parents  %>%
  select(ID_Clone, Trait, all_of(var)) %>%
  pivot_wider(names_from = Trait, values_from = all_of(var)) %>%
  mutate(
    N_Roots = 15 * N_Roots,
    FRY = 20 * FRY,
    ShY = 10 * ShY,
    DMC = 15 * DMC,
    StY = 10 * StY,
    Plant.Height = 5 * Plant.Height,
    HI = 10 * HI,
    StC = 10 * StC,
    Root.Le = 5 * Root.Le,
    Root.Di = 5 * Root.Di,
    Stem.D = 5 * Stem.D,
    Nstem.Plant = 5 * Nstem.Plant
  ) %>%
  mutate(pesos =
           rowSums(.[2:13], na.rm = TRUE))
}

pesos_BLUP <- calcular_pesos(selection_parents, "BLUP")
pesos_GEBV <- calcular_pesos(selection_parents, "GEBV")
pesos_GETGV <- calcular_pesos(selection_parents, "GETGV")
```


```{r pesos_save, include=FALSE} 
write.csv(selection_parents, "output/BLUPS_GEBV_GETG_all clones.csv" , row.names = FALSE)

write.csv(pesos_BLUP, "output/pesos_BLUPS_all clones.csv" , row.names = FALSE)

write.csv(pesos_GEBV, "output/pesos_GEBVS_all clones.csv" , row.names = FALSE)

write.csv(pesos_GETGV, "output/pesos_GETGVS_all clones.csv" , row.names = FALSE)
```

### Individual selection to each trait

```{r selection-functions, message=FALSE, warning=FALSE}
results_kappa <- data.frame()
SI <- c(10, 15, 20, 25, 30)

clones_sel_pesos <- function(pesos) {
  pesos %>%
    right_join(sel_parents) %>%
    droplevels() %>%
    arrange(desc(pesos)) %>%
    slice(1:(nlevels(ID_Clone) * (i / 100))) |>
    droplevels()
}

clone_sel_method <- function(data, method) {
  data %>%
    # Use a `mutate` para criar uma coluna temporária que armazena os valores de ordenação
    mutate(OrderingValue = ifelse(Weights == "acrescimo", get(method), -get(method))) %>%
    arrange(desc(OrderingValue)) %>%
    slice(1:(nlevels(ID_Clone) * (i / 100))) %>%
    droplevels() %>%
    select(-OrderingValue)  # Remova a coluna temporária
}

comb_sel <- function(var1, var2) {
  get(paste0("Clones_sel_", var1)) %>%
    full_join(get(paste0("Clones_sel_", var2))) %>%
    resca(BLUP, GEBV, GETGV, new_min = 0, new_max = 1) %>%
    mutate(!!paste0(var1, "_", var2) := (get(paste0(var1, "_res")) + get(paste0(var2, "_res"))) / 2) %>%
    arrange(desc(get(paste0(var1, "_", var2)))) %>%
    slice(1:nrow(Clones_sel_BLUP)) %>%
    droplevels()
}

calcular_media <- function(data, var) {
  data %>%
    filter(Trait == j) %>%
    select(all_of(var)) %>%
    summarise(mean(.[[1]], na.rm = T)) %>%
    pull()
}

calcular_media_sel <- function(data, var) {
  get(paste0("Clones_sel_", var)) %>%
    filter(Trait == j &
             ID_Clone %in% get(paste0("Clones_", var, "_sel"))$ID_Clone) %>%
    select(all_of(var)) %>%
    summarise(mean(.[[1]], na.rm = T)) %>%
    pull()
}

calcular_media_comb_sel <- function(data, var1, var2) {
  data %>%
    filter(Trait == j &
             ID_Clone %in% get(paste0("Comb_sel_", var1, "_", var2))$ID_Clone) %>%
    select(all_of(var1)) %>%
    summarise(mean(.[[1]], na.rm = T)) %>%
    pull()
}

# Função para calcular kappa
calcular_kappa <- function(var1, var2) {
  cohen.kappa(cbind(Clones_sel[[var1]], Clones_sel[[var2]]))[["kappa"]]
}
```


```{r loop-selection, message=FALSE, warning=FALSE}
# Melhore a clareza e a eficiência do loop de seleção
for (j in traits) {
  for (i in SI) {
    sel_parents <- droplevels(na.omit(subset(selection_parents, Trait == j)))
    
    # Aplicar a função clones_sel_pesos para selecionar clones
    Clones_GEBV_sel <- clones_sel_pesos(pesos_GEBV)
    Clones_GETGV_sel <- clones_sel_pesos(pesos_GETGV)
    Clones_BLUP_sel <- clones_sel_pesos(pesos_BLUP)
    
    # Aplicar a função clone_sel_method para métodos de seleção
    Clones_sel_BLUP <- clone_sel_method(sel_parents, "BLUP")
    Clones_sel_GEBV <- clone_sel_method(sel_parents, "GEBV")
    Clones_sel_GETGV <- clone_sel_method(sel_parents, "GETGV")
    
    # Calcular médias
    X0_BLUPS <- calcular_media(selection_parents, "BLUP")
    X0_GEBV <- calcular_media(selection_parents, "GEBV")
    X0_GETGV <- calcular_media(selection_parents, "GETGV")
    
    XS_BLUPS <- calcular_media_sel(Clones_sel_BLUP, "BLUP")
    XS_GEBV <- calcular_media_sel(Clones_sel_GEBV, "GEBV")
    XS_GETGV <- calcular_media_sel(Clones_sel_GETGV, "GETGV")
    
    # Combinar seleções
    Comb_sel_GEBV_BLUP <- comb_sel("GEBV", "BLUP")
    Comb_sel_GETGV_BLUP <- comb_sel("GETGV", "BLUP")
    Comb_sel_GETGV_GEBV <- comb_sel("GETGV", "GEBV")
    
    # Calcular médias combinadas
    XS_GEBV_BLUP <- calcular_media_comb_sel(selection_parents, "GEBV", "BLUP")
    XS_GETGV_BLUP <- calcular_media_comb_sel(selection_parents, "GETGV", "BLUP")
    XS_GETGV_GEBV <- calcular_media_comb_sel(selection_parents, "GETGV", "GEBV")
    
    # Selecionar clones
    Clones_sel <- transform(BLUPS, 
                           BLUPS_sel = as.integer(ID_Clone %in% Clones_sel_BLUP$ID_Clone),
                           GEBVS_sel = as.integer(ID_Clone %in% Clones_sel_GEBV$ID_Clone),
                           GETGV_sel = as.integer(ID_Clone %in% Clones_sel_GETGV$ID_Clone),
                           Comb_sel_GEBV_BLUP = as.integer(ID_Clone %in% Comb_sel_GEBV_BLUP$ID_Clone),
                           Comb_sel_GETGV_BLUP = as.integer(ID_Clone %in% Comb_sel_GETGV_BLUP$ID_Clone),
                           Comb_sel_GETGV_GEBV = as.integer(ID_Clone %in% Comb_sel_GETGV_GEBV$ID_Clone))

    # Calcular valores de kappa
    kappa_values <- data.frame(
      kappa_GEBV_BLUP = calcular_kappa("BLUPS_sel", "GEBVS_sel"),
      kappa_GETGV_BLUP = calcular_kappa("BLUPS_sel", "GETGV_sel"),
      kappa_GETGV_GEBV = calcular_kappa("GEBVS_sel", "GETGV_sel"),
      kappa_sel_GEBV_BLUP_BLUP = calcular_kappa("BLUPS_sel", "Comb_sel_GEBV_BLUP"),
      kappa_sel_GETGV_BLUP_BLUP = calcular_kappa("BLUPS_sel", "Comb_sel_GETGV_BLUP"),
      kappa_sel_GETGV_GEBV_BLUP = calcular_kappa("BLUPS_sel", "Comb_sel_GETGV_GEBV"),
      kappa_sel_GEBV_BLUP_GEBV = calcular_kappa("GEBVS_sel", "Comb_sel_GEBV_BLUP"),
      kappa_sel_GETGV_BLUP_GEBV = calcular_kappa("GEBVS_sel", "Comb_sel_GETGV_BLUP"),
      kappa_sel_GETGV_GEBV_GEBV = calcular_kappa("GEBVS_sel", "Comb_sel_GETGV_GEBV"),
      kappa_sel_GEBV_BLUP_GETGV = calcular_kappa("GETGV_sel", "Comb_sel_GEBV_BLUP"),
      kappa_sel_GETGV_BLUP_GETGV = calcular_kappa("GETGV_sel", "Comb_sel_GETGV_BLUP"),
      kappa_sel_GETGV_GEBV_GETGV = calcular_kappa("GETGV_sel", "Comb_sel_GETGV_GEBV")
    )
    
    # Coeficientes kappa
    coef_kappa <- data.frame(
      Trait = j,
      SI = i,
      X0 = media_pheno[[j]],
      X0_GEBV,
      X0_GETGV,
      X0_BLUPS,
      XS_BLUPS,
      XS_GEBV,
      XS_GETGV,
      XS_GEBV_BLUP,
      XS_GETGV_BLUP,
      XS_GETGV_GEBV
    )
    coef_kappa <- cbind(coef_kappa, kappa_values)
    
    # Anexar os resultados
    results_kappa <- rbind(results_kappa, coef_kappa)
  }
}
```


```{r selection_save, include=FALSE}
write.csv(results_kappa, "output/results_kappa.csv", row.names = FALSE)
```

Figure 3 Cohen’s Kappa of coincidence 

```{r kappa}
results_kappa |>
  pivot_longer(names_to = "Comparation",
               values_to = "Kappa",
               cols = 13:18) %>%
  ggplot(aes(x = Trait, y = SI, fill = Kappa)) +
  geom_tile() +
  facet_wrap(Comparation ~ ., labeller = as_labeller(
    c(
      kappa_GEBV_BLUP = "GEBV x BLUP",
      kappa_GETGV_BLUP = "GETGV x BLUP",
      kappa_GETGV_GEBV = "GEBV x GETGV",
      kappa_sel_GEBV_BLUP_BLUP = "GEBV_BLUP x BLUP",
      kappa_sel_GETGV_BLUP_BLUP = "GETGV_BLUP x BLUP",
      kappa_sel_GETGV_GEBV_BLUP = "GETGV_GEBV x BLUP"
    )
  ), ncol = 2) +
  scale_fill_viridis(discrete = FALSE, limits = c(-0.07, 1)) +
  labs(x = "" , y = "Selection Index", fill = "Kappa") +
  theme_bw() +
  theme(
    text = element_text(size = 20),
    legend.key.width = unit(1.5, 'cm'),
    legend.box = "horizontal",
    legend.position = "top",
    legend.background =  element_blank(),
    strip.background = element_blank(),
    panel.background = element_blank(),
    plot.background = element_blank(),
    axis.text.x = element_text(
      angle = 45,
      hjust = 1,
      vjust = 1
    )
  )
```


```{r kappa_save, include=FALSE}
ggsave(
  "output/kappa.tiff",
  width = 12,
  height = 12
)
```


Figure 4: Selection gains

```{r gains}
teste <-  results_kappa %>%
  select(Trait, SI, X0, XS_GEBV, XS_GETGV)

p <- list()

for (i in levels(factor(teste$Trait))) {
  breaks <- teste %>%
    filter(Trait == i) %>%
    droplevels() %>%
    group_by(Trait) %>%
    summarise(
      min_X0 = min(X0),
      max_XS = max(c(XS_GEBV, XS_GETGV)),
      mean_X0_XS = mean(c(min_X0, max_XS))
    ) %>%
    round_cols()
  
  p[[i]] <- teste %>%
    filter(Trait == i) %>%
    droplevels() %>%
    ggplot(aes(y = SI,
               x = start)) +
    geom_segment(
      aes(
        x = X0,
        xend = XS_GEBV,
        y = SI,
        yend = SI
      ),
      linewidth = 1,
      color = "gray80"
    ) +
    geom_point(
      data = teste %>%
        filter(Trait == i) %>%
        droplevels() %>%
        pivot_longer(
          names_to = "measure",
          values_to = "value",
          cols = c("X0", "XS_GEBV", "XS_GETGV")
        ),
      aes(y = SI,
          x = value,
          color = measure),
      size = 4,
      alpha = 0.75
    )  +
    scale_x_continuous(
      limits = ~ c(min(.x), max(.x)),
      breaks = c(breaks$min_X0, breaks$mean_X0_XS, breaks$max_XS),
      expand = expansion(mult = ifelse(i == "Plant.Height" , 0.25, 0.15))
    ) +
    scale_color_gdocs() +
    labs(x = i)  +
    theme(
      text = element_text(size = 15),
      legend.text = element_text(size = 15),
      legend.box = "horizontal",
      legend.direction = "horizontal",
      legend.position = "top",
      panel.spacing = unit(2, "lines"),
      legend.title = element_blank(),
      panel.background = element_blank(),
      panel.border = element_blank(),
      plot.background = element_blank(),
      legend.background = element_blank(),
      legend.box.background = element_blank(),
      legend.key = element_blank()
    )
}

annotate_figure(
  ggarrange(
    plotlist = p,
    nrow = 3,
    ncol = 4,
    common.legend = TRUE),
  left = text_grob("Selection Index", rot = 90, size = 20)
)
```


```{r gains_save, include=FALSE}
ggsave(
  "output/indice_selection_GEBV_GETGV.tiff",
  width = 12,
  height = 8,
  dpi = 300
)
```

Supplementary Table 3 - Cohen’s Kappa of coincidence 

```{r kappa_table}
results_kappa |>
  select(1,2, starts_with("kappa")) %>% 
  kbl(escape = F, align = 'c') |>
  kable_classic("hover", full_width = F, position = "center", fixed_thead = T)
```


```{r kappa_table_save, include=FALSE}
write.csv(results_kappa |>
  select( 1,2, starts_with("kappa")) , "output/results_values_kappa.csv", row.names = FALSE)
```

Supplementary Table 4 - Diferential selection


```{r diferencial_selection}
diferencial_selection <- results_kappa %>%
  mutate(DS_GEBV = ((XS_GEBV - X0) / X0)*100,
         DS_GETGV = ((XS_GETGV - X0) / X0)*100) %>%
  select(1:3, XS_GEBV, XS_GETGV, DS_GEBV, DS_GETGV)  

diferencial_selection %>%
  kbl(escape = F, align = 'c') |>
  kable_classic(
    "hover",
    full_width = F,
    position = "center",
    fixed_thead = T
  )
```


```{r diferencial_selection-save, include=FALSE}
write.csv(diferencial_selection, "output/dierencial_selecao.csv", row.names = FALSE)
```

### SNP-based heritability estimate 

SNP-based heritability estimate

```{r results_h2}
results_h2_GBLUP <- readRDS("output/results_cv_G_BLUP.RDS") %>%
  select(Trait, narrow_sense) %>% 
  group_by(Trait) %>%
  summarise(SNP_H2_narrow_sense = mean(narrow_sense))
```


```{r results_h2_save, include=FALSE}
write.csv(results_h2_GBLUP, "output/results_h2_GBLUP.csv", row.names = FALSE)
```

Table 2 Broad-sense heritability and SNP-based heritability

```{r}
H2 <- read.csv("output/H2_row_col_random.csv")
pheno_mean_sd <- read.csv("output/pheno_mean_sd.csv")

Broad_SNP_h2 <- H2 %>%
  rename(Trait = trait) %>% 
  full_join(results_h2_GBLUP) %>% 
  full_join(pheno_mean_sd %>%
              rename("Trait" = "variable")) %>%
  round_cols(digits = 2) %>%
  mutate(mean = str_c(mean, " (", min, " - ", max, ")")) %>%
  select(Trait, H2_Broad, H2_narrow, SNP_H2_narrow_sense, mean, cv)

Broad_SNP_h2 %>%
  kbl(escape = F, align = 'c') |>
  kable_classic(
    "hover",
    full_width = F,
    position = "center",
    fixed_thead = T
  )
```

```{r Broad_SNP_h2_save, include=FALSE}
write.csv(Broad_SNP_h2, "output/results_Broad_SNP_h2.csv", row.names = FALSE)
```
