---
title: "Genomic Selection"
author: 
  - Costa, W. G.^[Weverton Gomes da Costa, Pós-Doutorando, Embrapa Mandioca e Fruticultura, wevertonufv@gmail.com]
date: "`r Sys.Date()`"
site: workflowr::wflow_site
url: https://wevertongomescosta.github.io/Genomic-Selection-for-Drought-Tolerance-Using-Genome-Wide-SNPs-in-Casava/GWS
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
github-repo: wevertongomescosta/Genomic-Selection-for-Drought-Tolerance-Using-Genome-Wide-SNPs-in-Casava
---

```{r setup, include=F} 
knitr::opts_chunk$set(echo = T,
                      warning = F)
memory.limit(70000)
```

To perform the analyses, we will need the following packages:

```{r message=F}
library(readxl)
require(tidyverse)
library(kableExtra)
library(janitor)
library(genomicMateSelectR)
library(AGHmatrix)
require(ComplexHeatmap)
require(rrBLUP)
library(ggpmisc)
library(cvTools)
theme_set(theme_bw())
```

<br>

## Data

The data set is based in genotypes evalueted in five years (2016 to 2020), each year was considered as environment. 

### Names marker data

Primeiro vamos buscar os ID dos marcadores para cada clone.

```{r}
names <-
  read_excel("data/Phenotyping.xlsx", sheet = "GBS") %>%
  rename(Clone = `Names trials Petrolina`,
         ID_Clone = `Nome GBS`) %>%
  mutate(ID_Clone = str_replace_all(ID_Clone, ":", ".")) %>%
  select(Clone, ID_Clone)

names %>%
  head() %>%
  kbl(escape = F, align = 'c') %>%
  kable_classic(
    "hover",
    full_width = F,
    position = "center",
    fixed_thead = T
  )
```

### Phenotypic data

Agora vamos agrupar os IDs dos marcadores com os nomes dos clones.

```{r}
pheno <- readRDS("output/BLUPS.RDS") %>%
  inner_join(names) %>% # Join Phenotypic data with Genotypics names (ID_Clones) of the Clones
  mutate(Clone = as.factor(Clone),
         ID_Clone = as.factor(ID_Clone))

pheno %>%
  head() %>%
  kbl(escape = F, align = 'c') %>%
  kable_classic(
    "hover",
    full_width = F,
    position = "center",
    fixed_thead = T
  )
```

### Genotypic data

Agora vamos carregar os dados genotípicos dos marcadores GBS e corrigir os valores dos pares de base. Além disso, também vamos dividir a coluna alleles em duas colunas, para o alelo de referencia e o recessivo. E vamos selecionar as colunas com os nomes dos marcadores, alelos de referencia e as colunas com os IDs dos clones de acordo com os dados dos BLUPs.

```{r}
geno <- read.table("data/allchrAR08.txt", header = T)

geno <- geno %>%
  mutate_at(vars(12:ncol(geno)), funs(
    case_when(
      . == "A" ~ "AA",
      . == "R" ~ "AG",
      . == "W" ~ "AT",
      . == "M" ~ "AC",
      . == "C" ~ "CC",
      . == "S" ~ "CG",
      . == "Y" ~ "CT",
      . == "G" ~ "GG",
      . == "K" ~ "GT",
      . == "T" ~ "TT",
    )
  )) %>%
  separate(alleles, c("reference", "recess")) %>%
  select(rs, reference, recess, any_of(pheno$ID_Clone))

geno[1:5, 1:20] %>%
  kbl(escape = F, align = 'c') %>%
  kable_classic(
    "hover",
    full_width = F,
    position = "center",
    fixed_thead = T
  )
```

Agora precisamos fazer a conversão de pares de base para dosagem alélica de acordo com o alelo de referência. Também vou adcionar a coluna rs como nome das colunas. Depois vou excluir as colunas dos alelos de reference e recess.

```{r}
geno <- geno %>%
  mutate_at(vars(4:ncol(geno)), funs(case_when(
    . == paste(reference, reference, sep = "") ~ 2,
    . == paste(recess, recess, sep = "") ~ 0,
    T ~ 1
  ))) %>%
  column_to_rownames("rs") %>%
  select(-c(reference, recess))

geno[1:5, 1:15] %>%
  kbl(escape = F, align = 'c') %>%
  kable_classic(
    "hover",
    full_width = F,
    position = "center",
    fixed_thead = T
  )
```

Para converter no formato para realizar as análises de GWS temos que transpor a matriz de marcadores.

```{r}
geno <- geno %>%
  t()

geno[1:5, 1:5] %>%
  kbl(escape = F, align = 'c') %>%
  kable_classic(
    "hover",
    full_width = F,
    position = "center",
    fixed_thead = T
  )
```

Vamos verificar quantos clones apresentam dados genotipados com os marcadores.

```{r}
geno %>%
  dim() %>%
  t() %>%
  kbl(
    escape = F,
    align = 'c',
    col.names = c("Number of Clones", "Number of markers")
  ) %>%
  kable_classic(
    "hover",
    full_width = F,
    position = "center",
    fixed_thead = T
  )
```

Com a filtragem dos genótipos em comum, temos 414 genótipos e 27405 marcas.

Agora vamos filtrar os SNPS usando MAF de 0.01 e verificar quantos marcadores se manterão.

```{r}
geno <- maf_filter(geno, thresh = 0.01)

geno %>%
  dim() %>%
  t() %>%
  kbl(
    escape = F,
    align = 'c',
    col.names = c("Number of Clones", "Number of markers")
  ) %>%
  kable_classic(
    "hover",
    full_width = F,
    position = "center",
    fixed_thead = T
  )
```

Com o filtro de MAF a 1% restaram 22779 marcadores. Vou salvar a matriz agora para que possamos carregar ela caso seja necessário.

```{r}
saveRDS(geno, "data/SNPs.rds")
```

## Building the G matrix

Again, we will use the [AGHmatrix](https://cran.r-project.org/web/packages/AGHmatrix/vignettes/Tutorial_AGHmatrix.html) package [@amadeu_aghmatrix_2016] to build the G matrix:

```{r warning=F}
G_matrix <- Gmatrix(geno,
                    method = "VanRaden",
                    ploidy = 2,
                    missingValue = NA)
```

Now we have the whole G matrix (414 x 414), which we can represent using a heatmap:

```{r}
Heatmap(
  G_matrix,
  show_row_names = F,
  show_column_names = F,
  heatmap_legend_param = list(title = "Res")
)
```

"Res" in the heatmap legend title is for "Resemblance".

## RRBLUP

The Ridge Regression BLUP, or RRBLUP, will predict the marker effect. In the RRBLUP, we will use the matrix of markers directly. For this purpose, we will use only individuals with BLUps and SNPs available. 

```{r}
pheno <- pheno %>%
  filter(ID_Clone %in% rownames(geno)) %>%
  droplevels()
```

For this purpose, we will use the [rrBLUP](https://cran.r-project.org/web/packages/rrBLUP/rrBLUP.pdf) package [@endelman_2011]. In the code below, *y* is for the vector with the means, *Z* is where we will insert the SNPs matrix, *K* is for a covariance matrix for the random effects, which will be and identity matrix by default; and *X* is a design matrix for the fixed effects, which will be a vector of ones (**1**) by default. Note that we are returning to the "1, 0, -1" codification in the SNPs matrix. This is a requirement of the _rrBlUP_ package.

```{r}
result_RR_BLUP <- data.frame(
  Clone = unique(pheno$Clone),
  ID_Clone = unique(pheno$ID_Clone),
  stringsAsFactors = F
)

traits <- colnames(pheno)[2:19]

for (i in traits) {
  RRBLUP <- mixed.solve(y = pheno[[i]],
                        Z = geno - 1,
                        K = NULL,
                        X = NULL)
  GEBV <- geno %*% RRBLUP$u
  result <- data.frame(ID_Clone = rownames(GEBV),
                       stringsAsFactors = F)
  result[, i] <- data.frame(GEBV)
  result_RR_BLUP <-
    merge(result_RR_BLUP, result, by = "ID_Clone", all.x = T)
}
```

### GEBV x BLUP

The scatter plot above represents the additive genetic value of each marker. Once we have acquired these values, we may calculate the Genomic Estimated Breeding Values (GEBV) of the genotypes. Aqui, iremos adicionar os valores médios dos valores fenotípicos aos GEBVs obtidos pelo RR-BLUP e aos BLUPs para uma melhor comparação. These are the product of the SNPs matrix with the vector of the markers' genetic values:

```{r}
media_pheno <- read.table("output/media_pheno.csv")

colnames(media_pheno) <-
  c(colnames(result_RR_BLUP[3:ncol(result_RR_BLUP)]))

for (i in traits) {
  result_RR_BLUP[i] <- result_RR_BLUP[i] + media_pheno[, i]
}

GEBVS_RR_BLUP <- result_RR_BLUP %>%
  pivot_longer(NR.P:Staygreen,
               names_to = "Variable",
               values_to = "GEBVS")

BLUPS_MED <- pheno

for (i in traits) {
  BLUPS_MED[i] <- BLUPS_MED[i] + media_pheno[, i]
}

BLUPS_MED <- BLUPS_MED %>%
  pivot_longer(NR.P:Staygreen,
               names_to = "Variable",
               values_to = "BLUPS")

data_gws_RR_BLUP <- GEBVS_RR_BLUP %>%
  full_join(BLUPS_MED)
```

Agora vamos produzir os gráficos para cada variável comparando a correlação entre os GEBVs e BLUPs.

```{r}
data_gws_RR_BLUP %>%
  ggplot(aes(x = BLUPS, y = GEBVS)) +
  geom_point(aes(color = GEBVS), show.legend = F) +
  geom_smooth(method = lm, se = F) +
  stat_poly_eq(formula = y ~ x, label.y = "bottom") +
  scale_color_gradient(low = '#c80000', high = '#FFFF00') +
  facet_wrap(. ~ Variable, ncol = 6, scales = "free") +
  expand_limits(y = 0, x = 0) +
  ggtitle("RR-BLUP")
```

Agora vamos comparar os modelos por meio de um boxplot.

```{r}
data_gws_RR_BLUP %>%
  pivot_longer(4:5, names_to = "Method", values_to = "Values") %>%
  ggplot(aes(x = , y = Values, fill = Method)) +
  geom_boxplot() +
  facet_wrap(. ~ Variable, ncol = 6, scales = "free")  +
  expand_limits(y = 0, x = 0) +
  ggtitle("RR-BLUP")
```

<br>

## GBLUP

In the GBLUP, we will use the G matrix instead of the SNPs matrix. Thus, we will obtain the GEBV directly. Note that we will need to build the G matrix again, since some genotypes were dropped after our filtering. The rrBLUP package has a function called "A.mat" that build the Additive Genomic Matrix from a SNP matrix with "-1,0,1" codification:

```{r}
result_G_BLUP <- data.frame(
  Clone = unique(pheno$Clone),
  ID_Clone = unique(pheno$ID_Clone),
  stringsAsFactors = F
)

for (i in traits) {
  GBLUP <- mixed.solve(pheno[[i]], K = A.mat(geno - 1))
  result <- data.frame(ID_Clone = rownames(GBLUP$u),
                       stringsAsFactors = F)
  result[, i] <- data.frame(GBLUP$u)
  result_G_BLUP <-
    merge(result_G_BLUP, result, by = "ID_Clone", all.x = T)
}
```

Aqui, iremos adicionar novamente os valores médios dos valores fenotípicos aos GEBVs obtidos pelo G-BLUP.

```{r}
for(i in traits) {
  result_G_BLUP[i] <- result_G_BLUP[i] + media_pheno[, i]
}

GEBVS_G_BLUP <- result_G_BLUP %>%
  pivot_longer(NR.P:Staygreen,
               names_to = "Variable",
               values_to = "GEBVS")

data_gws_G_BLUP <- GEBVS_G_BLUP %>%
  full_join(BLUPS_MED)
```

Agora vamos produzir os gráficos para cada variável comparando a correlação entre os GEBVs e BLUPs.

```{r}
data_gws_G_BLUP %>%
  ggplot(aes(x = BLUPS, y = GEBVS)) +
  geom_point(aes(color = GEBVS), show.legend = F) +
  geom_smooth(method = lm, se = F) +
  stat_poly_eq(formula = y ~ x, label.y = "bottom") +
  scale_color_gradient(low = '#c80000', high = '#FFFF00') +
  facet_wrap(. ~ Variable, ncol = 6, scales = "free") +
  expand_limits(y = 0, x = 0) +
  ggtitle("G-BLUP")
```

Agora vamos comparar os modelos por meio de um boxplot.

```{r}
data_gws_G_BLUP %>%
  pivot_longer(4:5, names_to = "Method", values_to = "Values") %>%
  ggplot(aes(x = , y = Values, fill = Method)) +
  geom_boxplot() +
  facet_wrap(. ~ Variable, ncol = 6, scales = "free")  +
  expand_limits(y = 0, x = 0) +
  ggtitle("G-BLUP")
```

Agora vamos produzir os gráficos para cada variável comparando a correlação entre os GEBVs do RR-BLUP e GEBVs do G-BLUP.

```{r}
GEBVS_RR_BLUP <- GEBVS_RR_BLUP %>%
  rename(GEBVS_RR_BLUP = GEBVS)

data_gws <- GEBVS_RR_BLUP %>%
  full_join(GEBVS_G_BLUP)

data_gws %>%
  ggplot(aes(x = GEBVS_RR_BLUP, y = GEBVS)) +
  geom_point() +
  geom_smooth(method = lm, se = F) +
  stat_poly_eq(formula = y ~ x, label.y = "bottom") +
  facet_wrap(. ~ Variable, ncol = 6, scales = "free") +
  expand_limits(y = 0, x = 0) +
  ggtitle("G-BLUP")
```


## Cross-validation

To prove that the prediction is accurate, we should perform a cross-validation (CV) scheme. For this purpose, we divide the data into a training set and a validation set. 

First we separate the data into *k* folds. Then, we attribute *NA* for one fold and try to predict the data from this fold based on the others. When selecting the number of folds, one must prioritize the balance between the number of observations in each fold. 

In addition, this process should be repeated for further validation. The step-by-step below will guide the CV in the data we are analysing.

###  1. Determine the number of folds and repetitions

```{r}
nfolds = 5
nrept = 5
```

Since we defined 5 folds, our data will be divided into 5 parts with 83 observations each.

### 2. Match the order of the data and the rows of the SNP matrix

The order is decreasing or increasing (numeric or alphabetical) regarding the name of the genotypes.

```{r}
pheno <- pheno[order(pheno$ID_Clone, decreasing = F),]
geno <- geno[order(row.names(geno)),]
all(rownames(geno) == pheno$ID_Clone)
```

### 3. Add a column indicating a number for each observation

This will be useful to assign each observation for a fold, which will be the next step.

```{r}
pheno$ID <- factor(1:nrow(pheno))
```

### 4. Folds assignment

In this step, we will assign each observation to a fold. Bear in mind that for each repetition, the folds will comprise different observations. The purpose of the repetition is to make sure of the randomness of the assignment step. 

In this step, we will use the [cvTools](https://cran.r-project.org/web/packages/cvTools/cvTools.pdf) package [@cvTools]

```{r}
set.seed(100)

sort <- list()

for (a in 1:nrept) {
  for (j in 1:nfolds) {
    folds <- cvFolds(nlevels(pheno$ID),
                     type = "random",
                     K = 5,
                     R = 1)
    Sample <- cbind(folds$which, folds$subsets)
    cv <- split(Sample[, 2], f = Sample[, 1])
  }
  sort[[a]] <- cv
}

rm(a, folds, j, cv, Sample)
```

### 5. Cross-validation

The next step is the very CV. Here, we will define the linear predictor and the lists that will be useful in the loop. The first list, here called "fold.list", contains the folds assignation that we built in the previous step. The second ("results_cv") is empty and will store the outputs of each iteration of the loop.

```{r}
fold.list <- sort
results <- list()
results_cv <- data.frame()
```

Then, we will construct the loop. Each iteration will assign *NA* for a different fold, and we will use the other folds to predict the missing values. Note that the folds vary for each repetition.

```{r}
for(j in traits) {
  for (z in 1:length(fold.list)) {
    for (i in 1:nfolds) {
      # Training set
      train_data <- pheno
      
      # Validation set
      train_data[train_data$ID %in% fold.list[[z]][[i]], j] <- NA
      
      # Fitting model
      GBLUP <- mixed.solve(train_data[[j]], K = A.mat(geno - 1))
      
      # GEBV
      Pred <- data.frame(Yhat = GBLUP$u, G = pheno$ID)
      
      rownames(Pred) <- rownames(geno)
      
      # Predicted GEBV
      results[[i]] <- Pred[Pred[, "G"] %in% fold.list[[z]][[i]], ]
      
      # Remove
      #rm(GBLUP, Pred, train_data)
    }
    
    GEBV <- do.call(rbind, results)
    GEBV <- GEBV[order(GEBV$G), ]
    
    # Log
    log <- all(GEBV$G == pheno$ID)
    
    # Results
    result_cv <- data.frame(
      Trait = j,
      Rep = z,
      Log = log,
      Ac = round(cor(GEBV$Yhat, train_data[[j]], use = "na.or.complete"), 3),
      MSPE = round(mean(((GEBV$Yhat - train_data[[j]]) ^ 2
      ), na.rm = T), 3)
    )
    
    results_cv <-
      rbind(results_cv, result_cv)
  }
}

results_cv %>%
  kbl(escape = F, align = 'c') %>%
  kable_classic(
    "hover",
    full_width = F,
    position = "center",
    fixed_thead = T
    
  )

write.csv(results_cv, "output/results_cv.csv")
```

The object "result_cv" is divided by repetition. In the "result_cv" objects for each repetition, "Rep" is the number of the repetition, "Log" is a diagnostic indicating if the order of the predicted breeding values matches the order of the adjusted means, "Ac" is the prediction accuracy (correlation between the GEBV and adjusted means), and "MSPE" is the mean square prediction error (the lower, the better).

Agora vamos plotar os resultados de acurácias para cada característica

```{r}
results_cv %>%
  group_by(Trait) %>%
  summarise(
    mean.Ac = mean(Ac),
    sd.Ac = sd(Ac),
    mean.MSPE = mean(MSPE),
    sd.MSPE = sd(MSPE)
  ) %>%
  ggplot(aes(x = Trait, y = mean.Ac , fill = Trait)) +
  geom_col(alpha = 0.8,
           width = 0.85,
           show.legend = F) +
  geom_errorbar(
    aes(ymin = mean.Ac - sd.Ac, ymax = mean.Ac + sd.Ac),
    width = .2,
    position = position_dodge(.9)
  ) +
  labs(y = "Accuracy")
```
